Design and Implementation of Probabilistic Programming Language Anglican

### 2017-02-25 paper/paper.tex
wip commit

\newtoggle{full}
\input{toggle}

@@

Probabilistic programming systems \cite{GMR+08,MSP14,WVM14,GS15}
represent generative models as programs written in a specialized
language that provides syntax for the definition and conditioning of
random variables. The code for such models is generally concise,
modular, and easy to modify or extend. Typically inference can be
performed for any probabilistic program using one or more generic
inference techniques provided by the system backend, such as
Metropolis-Hastings \cite{WSG11,MSP14,YHG14}, Hamiltonian Monte Carlo
\cite{SDT14}, expectation propagation \cite{MWG+10}, and extensions
of Sequential Monte Carlo \cite{WVM14,MYM+15,PWD+14} methods.
These generic techniques may be less statistically efficient
than techniques tailored to a specific model. However,
probabilistic programming facilitates simpler implementations of
models making the inference inherently faster.

@@

    ;;; Observe samples from the distribution.
    (loop [observations data]
      (when (not-empty observations)
        ;; Retrieve the first observation as `o',
        ;; and store the rest of observations in 
        ;; 'observations*'.
        (let [[o & observations*] observations]
          ;; Observe 'o' from the guessed 
          ;; distribution 'd'.
          (observe d o))
        ;; Proceed to the next iteration with
        ;; the rest of observations.
        (recur observations*)))

    ;;; Return the distribution and parameters.
    [d a b]))
\end{lstlisting}
%\end{minipage}
The query builds a model for the input $\texttt{data}$, 
a sequence of data points.
It defines a probability distribution on three variables, 
$\texttt{d} \in \{\texttt{normal},\texttt{gamma}\}$
for a distribution type, and $\texttt{a}$ and $\texttt{b}$ for
positive parameters for the type. Concretely,
using the $\texttt{sample}$ forms, the query
first defines a so called prior distribution on these three variables,
and then it adjusts this prior distribution based on observations in
$\texttt{data}$ using the \texttt{observe} form.
Samples from this adapted distribution (also called posterior distribution)
can be obtained by running the query under one of Anglican's
inference algorithms.

Internally, an Anglican query is represented by a computation in
\textit{continuation passing style} (CPS)~\cite{AJ89}, and inference algorithms
exploit the CPS structure of the code to intercept probabilistic
operations in an algorithm-specific way\footnote{\cite{GS15}
also describe a CPS-based implementation of a probabilistic
programming language.}. Among the available
inference algorithms there are Particle Cascade~\cite{PWD+14},
Lightweight Metropolis-Hastings~\cite{WSG11}, Iterative Conditional
Sequential Monte-Carlo (Particle Gibbs)~\cite{WVM14}, and others.
Inference on Anglican queries generates a lazy sequence of
samples, which can be processed asynchronously in Clojure
code for analysis, integration, and decision making.

Clojure (and Anglican) runs on the JVM and gets access to a wide choice of Java
libraries for data processing, networking, presentation, and imaging.
Conversely, Anglican queries can be called from Java and other JVM languages.
Programs involving Anglican queries can be deployed as JVM
\textit{jars}, and run without modification on any platform for
which JVM is available.

A probabilistic program, or query, mostly runs deterministic
code, except for certain checkpoints, in which probabilities are
involved, and normal, linear execution of the program is
disrupted.
\iftoggle{full}{%
In Anglican and similar languages there are two types
of such checkpoints:

\begin{itemize}
    \item drawing a value from a random source (\texttt{sample});
    \item conditioning a computed value on a random source
        (\texttt{observe}).
\end{itemize}

}{%
}%
Anglican can be mostly implemented as a regular programming
language, except for the handling of these checkpoints.
\iftoggle{full}{%
Depending on the \textit{inference algorithm}, \texttt{sample}
and \texttt{observe} may result in implicit input/output
operations and control changes. For example, \texttt{observe} in
particle filtering inference algorithms \cite{WVM14} is a non-deterministic
control statement at which a particle (corresponding to
a user-level thread executing a program)
can be either replicated or terminated. Similarly, in 
Metropolis-Hastings \cite{WSG11},
\texttt{sample} is both an input
statement which `reads' values from a random source,
and a non-deterministic control statement
(with delayed effect), eventually affecting acceptance or
rejection of a sample.

>>>>

\newtoggle{full}
\input{toggle-ifl }

@@

Probabilistic programming systems \cite{GMR+08,MSP14,WVM14,GS15}
represent generative models as programs written in a specialized
language that provides syntax for the definition and conditioning of
random variables. The code for such models is generally concise,
modular, and easy to modify or extend. Typically inference can be
performed for any probabilistic program using one or more generic
inference techniques provided by the system backend, such as
Metropolis-Hastings \cite{WSG11,MSP14,YHG14}, Hamiltonian Monte Carlo
\cite{SDT14}, expectation propagation \cite{MWG+10}, and extensions
of Sequential Monte Carlo \cite{WVM14,MYM+15,PWD+14} methods.
These generic techniques may be less statistically efficient
than techniques tailored to a specific model. However,
probabilistic programming facilitates simpler implementations of
models making the inference inherently faster.
\jw{I'm not sure I understand how PP facilitates simpler implmenetations that make inference inherently faster.}

@@

    ;;; Observe samples from the distribution.
    (loop [observations data]
      (when (not-empty observations)
        ;; Retrieve the first observation as `o'
        (let [o (first observations)]
          ;; Observe 'o' from the guessed 
          ;; distribution 'd'.
          (observe d o))
        ;; Proceed to the next iteration with
        ;; the rest of observations.
        (recur (rest observations))))

    ;;; Return the distribution and parameters.
    [d a b]))
\end{lstlisting}

%\end{minipage}
The query builds a model for the input $\texttt{data}$, 
a sequence of data points.
It defines a probability distribution on three variables, 
$\texttt{d} \in \{\texttt{normal},\texttt{gamma}\}$
for a distribution type, and $\texttt{a}$ and $\texttt{b}$ for
positive parameters for the type. Concretely,
using the $\texttt{sample}$ forms, the query
first defines a so called prior distribution on these three variables,
and then it adjusts this prior distribution based on observations in
$\texttt{data}$ using the \texttt{observe} form.
Samples from this conditioned distribution (also called posterior distribution)
can be obtained by running the query under one of Anglican's
inference algorithms.

Clojure (and Anglican) programs run on the JVM and are able to make
use of a wide range of Java libraries for data processing,
networking, presentation, and imaging. Conversely, Anglican queries
can be called from Java and other JVM languages. Programs involving
Anglican queries can be deployed as JVM \textit{jars}, and run without modification on any platform for which the JVM is available.

A probabilistic program, or query, mostly runs deterministic code,
except for \texttt{sample} and \texttt{observe} forms, which are
probabilistic in nature. For this reason, Anglican can be mostly implemented as a regular programming language, except for the handling of these two special forms, which we refer to as checkpoints. Internally, an Anglican query is represented by a
computation in \textit{continuation passing style} (CPS)~\cite{AJ89}.
Inference algorithms exploit the CPS structure of the code to
evaluate \texttt{sample} and \texttt{observe} checkpoints in an
algorithm-specific manner\footnote{\cite{GS15} also describe a
CPS-based implementation of a probabilistic programming language.}.
The available inference algorithms include the Particle
Cascade~\cite{PWD+14}, Lightweight Metropolis-Hastings~\cite{WSG11},
Iterative Conditional Sequential Monte-Carlo (Particle Gibbs)
\cite{WVM14}, and others. Inference on Anglican queries generates a
lazy sequence of samples, which can be processed asynchronously in
Clojure for analysis, integration, and decision making.

% \iftoggle{full}{%
% In Anglican and similar languages there are two types
% of such checkpoints:

% \begin{itemize}
%     \item drawing a value from a random source (\texttt{sample});
%     \item conditioning a computed value on a random source
%         (\texttt{observe}).
% \end{itemize}

% }{%
% }%
\iftoggle{full}{%
Depending on the \textit{inference algorithm}, \texttt{sample}
and \texttt{observe} may result in implicit input/output
operations and control changes. For example, \texttt{observe} in
particle filtering inference algorithms \cite{WVM14} is a non-deterministic
control statement at which a particle (corresponding to
a user-level thread executing a program)
can be either replicated or terminated. Similarly, in 
Metropolis-Hastings \cite{WSG11},
\texttt{sample} is both an input
statement which `reads' values from a random source,
and a non-deterministic control statement
(with delayed effect), eventually affecting acceptance or
rejection of a sample.
